Project Description -- Erosion, Transport, Anomalous Diffusion, and Collapse in Granular and Porous Media

### 2019-11-12 docs/proposals/NSF/ProjDesc/ProjDesc.tex
NM

	These filtration processes, however, rely on an understanding of complex transport phenomenon in porous media. Further, the action of continually flowing groundwater can alter medium properties through erosive and transportive processes. Thus, it becomes necessary to understand the transport of tracers through complex media whose properties are changing {\em dynamically} in response to the intervening fluid flow.

	This proposal aims to develop a suite of computational tools geared towards developing a deeper understanding of the physical processes by which groundwater flow can alter porous-media properties and to characterize the associated changes in transport within the evolving medium, e.g.~anomalous dispersion. A recognized complexity of naturally occurring media is that their macroscopic properties, e.g.~permeability and mechanical strength, can be highly heterogeneous and anisotropic. These properties can change over short timescales, for example sinkhole collapse or hydraulic fracturing, or over long timescales, for example the slow erosion or dissolution of solid material. Potential exists for feedback between these timescales: for example the slow wearing process reaching a critical threshold beyond which catastrophic events become more likely.

@@

In particular, we view the slow processes of erosion and dissolution as instrumental in altering porous-medium properties. First, by deteriorating solid material, erosion alters the size distribution of grains. This modified distribution can in turn affect transport by enabling smaller grains to be sieved through networks of larger ones. A second and more subtle way that erosion can alter medium properties is by changing the {\em shape} distribution of grains. For instance, erosion can carve grains into slender shapes aligned primarily in the direction of the dominant flow, thereby creating {\em anisotropic} characteristics. Likewise, chemical dissolution can carve well-defined conduit paths in karst networks which strongly alter transport properties \cite{szymczak2009wormhole}.

@@

Only recently have these ideas been extended to low Reynolds number flows  \cite{mitchell2017generalized, Quaife2018} --- the regime applicable to porous media. While the governing fluid-flow equations become linear in this regime, porous-media applications present a host of new challenges, for example:
\begin{itemize}
\item Porous media typically involves dense packings so that close-range effects and even nearly contacting bodies must be considered.
\item The Stokes limit introduces {\em longer} range effects than the inertial limit does, thus coupling the shape evolution of {\em all} grains and conduits to a higher degree.
\item Somewhat counterintuitively, erosion does not always act to smooth shapes but rather can create sharp features and corners \cite{Ristroph2012}. These features can be difficult to resolve and, further, can trigger numerical instabilities \cite{Quaife2018}. Dissolution, meanwhile, can initiate a roughening transition \cite{claudin2017dissolution}, leading to similar numerical difficulties.
\item Additionally effects, such as sedimentation, transport, and compaction constantly work to reconfigure the medium as its constituents and boundaries degrade.
\end{itemize}
These challenges require innovative computational techniques, many of which have been developed recently \cite{Quaife2018} and many of which await resolution. An ultimate goal of this proposal is to achieve high-fidelity simulations of $O(100-1000)$ mobile bodies undergoing simultaneous erosion, transport, sedimentation, and compaction, with the computational efficiency to run large ensembles for statistical analysis.
Fortunately, some closely related laboratory experiments on erosion in colloidal systems are already being explored in the Datta Lab at Princeton \cite{bizmark2019multiscale}. The experiments offer a wonderful opportunity to guide, calibrate, and validate theoretical developments.

While erosion and dissolution lead to gradual changes in a medium, other events like sinkhole collapse can occur suddenly and catastrophically. It is likely that all of the mechanisms listed above --- erosion, dissolution, transport, compaction, and gravitational collapse --- contribute to sinkhole formation, with the slowest acting mechanisms creating conditions favorable for the final act of gravity. Sinkhole formation has been explored with preliminary experiments run at the Geophysical Fluid Dynamics Institute (GFDI) at Florida State University \cite{tao2014experimental}. These scaled-down experiments offer the ability to probe various mechanisms behind sinkhole formation, for example a difference in hydraulic head between the confined and unconfined aquifers. An ultimate goal of this project is to merge these two lines of inquiry, on erosive processes and sinkhole formation, to gain a deeper understanding the influence of mechanically or chemically induced erosion on gravitational collapse. Ultimately, we envision the high-fidelity erosion simulations to inform coarse-grained models for transport and dynamics of macroscopic medium properties. These parameterized models could then be combined with well-developed multi-phase models \tocite to be used for the quantitative description and prediction of sinkhole collapse.

@@

% Other stuff that could be included: McCurdy and Wang work, Multiphase work

>>>>

	An understanding of these filtration and contamination processes relies on the complex phenomenon of transport in porous media. Further, the action of continually flowing groundwater can alter medium properties through erosive and transportive processes. Thus, it becomes necessary to understand transport through complex media whose properties change {\em dynamically} in response to the intervening fluid flow. Such effects become most prominent during rapid events, like the gravitational collapse of a sinkhole, during which medium properties can change suddenly and dramatically. Though less obvious, the accumulation of slower process, such as mechanical or chemical erosion, can also substantially alter medium properties.

	This proposal aims to develop a suite of computational tools geared towards developing a deeper understanding of the physical processes by which groundwater flow can alter porous-media properties and to characterize the associated changes in transport within the evolving medium, for example the anomalous dispersion of tracers. A recognized complexity of naturally occurring media is the heterogeneity and anisotropy of macroscopic properties such as permeability and mechanical strength. These properties can change over short timescales, for example sinkhole collapse or hydraulic fracturing, or over long timescales, for example the slow erosion or dissolution of solid material. Potential exists for feedback between these timescales: for example the slow wearing process reaching a critical threshold beyond which catastrophic events become more likely.

@@

In particular, we view the accumulated effects of slow processes like erosion and dissolution as instrumental in altering porous-medium properties. First, by deteriorating solid material, erosion alters the size distribution of grains, enabling smaller grains to be sieved through networks of larger ones. A second and more subtle way that erosion can alter medium properties is by changing the {\em shape} distribution of grains. For instance, erosion can carve grains into slender shapes aligned primarily in the direction of the dominant flow, thereby creating {\em anisotropic} characteristics. Likewise, chemical dissolution can carve well-defined conduit paths in karst networks which strongly alter transport properties \cite{szymczak2009wormhole}.

@@

Only recently have these ideas been extended to the low-Reynolds-number regime that is applicable to porous media  \cite{mitchell2017generalized, Quaife2018}. While the governing fluid-flow equations become linear in this regime, porous-media applications present a host of new challenges, for example:
\begin{itemize}
\item Porous media typically involves dense packings so that close-range effects and even nearly contacting bodies must be considered.
\item The Stokes limit introduces {\em longer} range effects than the inertial limit does, thus coupling the shape evolution of {\em all} grains and conduits to a higher degree.
\item Somewhat counterintuitively, erosion does not always act to smooth shapes but rather can create sharp features and corners \cite{Ristroph2012}. These features can be difficult to resolve and, further, can trigger numerical instabilities \cite{Quaife2018}. Dissolution, meanwhile, can initiate a roughening transition \cite{claudin2017dissolution}, leading to similar numerical difficulties.
\item Additionally effects, such as sedimentation, transport, and compaction constantly work to reconfigure the medium as its constituents and boundaries degrade.
\end{itemize}
These challenges require innovative computational techniques, some of which have been developed recently \cite{Quaife2018} and some of which await resolution. An ultimate goal of this proposal is to achieve high-fidelity simulations of $O(100-1000)$ mobile bodies undergoing simultaneous erosion, transport, sedimentation, and compaction, with the computational efficiency required to run large ensembles for statistical analysis.
Fortunately, some closely related laboratory experiments on erosion in colloidal systems are already being explored in the Datta Lab at Princeton \cite{bizmark2019multiscale}. The experiments offer a wonderful opportunity to guide, calibrate, and validate theoretical developments, as do proposed experiments at the Geophysical Fluid Dynamics Institute.

While erosion and dissolution lead to gradual changes in a medium, other events like sinkhole collapse can occur suddenly and catastrophically. Most likely, all of the mechanisms listed above --- erosion, dissolution, transport, compaction, and gravitational collapse --- contribute to sinkhole formation, with the slowest acting mechanisms creating conditions favorable for the final act of gravity. Sinkhole formation has been explored with preliminary experiments run at the Geophysical Fluid Dynamics Institute (GFDI) at Florida State University \cite{tao2014experimental}. These scaled-down experiments offer the ability to probe various mechanisms behind sinkhole formation, for example a difference in hydraulic head between the confined and unconfined aquifers. An ultimate goal of this project is to merge these two lines of inquiry to gain a deeper understanding the influence of mechanically or chemically induced erosion on gravitational collapse. Ultimately, we envision the high-fidelity erosion simulations to inform coarse-grained models for: (i) transport and dispersion; and (ii) dynamic evolution of macroscopic medium properties. These parameterized models could then be combined with well-developed multi-phase models \cite{Imma2019, eastham2019multiphase} to be used for the quantitative description and prediction of sinkhole collapse.

@@
